Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics

### 2014-10-02 anderson-etal-blackswan-timeseries.tex
Switch to markdown ms; add to ms

@@

\noindent
\textit{Authors are listed alphabetically (ABCD!) and open to reordering.}

\noindent
\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
\noindent
\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

\noindent
\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada

\noindent
\textsuperscript{*}Corresponding author: Sean C. Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca

\clearpage

\setcounter{page}{1}

\noindent
\textit{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.

%When they do occur they tend to be driven by ... little evidence....

\section{Abstract}

Black swans are statistically improbable events that have profound implications when they occur. Such events are known to occur in financial, social, and environmental systems; the magnitude of environmental extremes is projected to increase with climate change. But, the prevalence and magnitude of black swans in ecological population dynamics is unknown. Here, we develop a probability model that allows us to estimate the degree of heavy-tailedness in ecological process noise. We apply our model to \NPops\ time series from around the world across \NOrders\ taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations and most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). When they occur, they tend to be driven by climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these elements. We find little evidence of intrinsic life-history correlates.
Our results suggest [FOR ECOLOGICAL MODELLING]
and the importance of establishing management strategies that are robust to extreme events \ldots.

%lso mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information

%- conclude with ecological, modelling, policy, and management implications

\section{Introduction}

\begin{enumerate}
\item Black swans are statistically improbable events whose occurrence has major ramifications. One of the most profound black swans in ecology was the asteroid marking the mass extinctions at the K-T boundary. Today, it is the extremeness of climate that is expected to cause the greatest societal damage. While extremes in the physical environment are present and widely understood, it remains unclear the extent to which ecological systems buffer or suffer from black swans.

\item What is the existing evidence for heavy tails in ecological time series?

\item There are two main possible reasons for why we may find little evidence of ecological black swans. (1) don't exist, (2) a paucity of appropriate methods or length/quality of time series.

\item Here we develop \ldots our research questions and main conclusions.

\end{enumerate}

%Increasing realization about the importance of extreme events in the environment (REFs)
%and of the importance of ecological surprises and black swans (REFs)

\bigskip
Papers to work in:

\citep{inchausti2002,halley2002,inchausti2001}

\citep{jentsch2007}

\citep{ward2007}

\citep{garcia-carreras2011}
\citep{sornette2009}

\citep{nunez2012}

\citep{thompson2013}
\citep{beaugrand2012}
\citep{pine-iii2009}

\citep{doak2008}

\citep{smale2013}

\citep{easterling2000}
\citep{scheffer2003}
\citep{katz2005}

\citep{taleb2007}

\citep{vasseur2014}

\citep{vert-pre2013}
\citep{lindenmayer2010}

\citep{valpine2002}
\citep{gregory2010}
\citet{garcia-carreras2011}
\citet{brook2006}
\citep{herrandoprez2014}

\citep{sibly2005, ziebarth2010}

\ldots

\section{Methods}

\begin{enumerate}
  \item Short overview paragraph
  \item The data: global population dynamics database; briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).
  \item The heavy-tailed Gompertz population model. Mention the simulation testing but don't go into details here.
  \item Mention the alternative population models we fit.
  \item Modelling the possible correlates (mixed effects models with binomial or beta distributions and logit links; nested random effects for taxonomic class and taxonomic order). Life-history data from \citet{brook2006a}.
  \end{enumerate}

The Student-t Gompertz model is:
\begin{align*}
x_t &= \ln N_t\\
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$. The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$. The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$. If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.

We fit alternative models that allowed for autocorrelation of the residuals, allowed for observation error, and allowed the functional form of the population dynamics to be represented as Ricker-logistic (Supporting Material).

%Mention that we simulation tested the model.

%Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.

%$\nu$ prior: \citep{fernandez1998}
%$\sigma$ prior: \citep{gelman2006c}
%consider alternative inverse prior on $\nu$ as in \citep{gelman2014}

%justify why we only present these as hypotheses but do not formally model them

\bigskip
Supporting material to be noted:

\begin{enumerate}

  \item Table showing the taxonomic breakdown, number of interpolated points, etc (Table~S1).

  \item Plot showing time series we assumed were recorded as log transformed  (Fig.~\ref{fig:log10-assumed}).

  \item Plot showing all the time series we used with interpolated and zero imputed values (Fig.~\ref{fig:all-ts}).

  \item Plot showing the priors (Fig.~\ref{fig:priors}).
\end{enumerate}

\section{Results}

Main figures and tables:

\begin{enumerate}
  \item Fig.~\ref{fig:didactic} illustrates the method: it shows the t-distribution tails, example simulated time series, and model fits to those time series.

  \item Fig.~\ref{fig:nu-coefs} shows the posterior distributions of the estimated $\nu$ values. These are split by taxonomic class and order.

  \item Fig.~\ref{fig:correlates} shows possible biological and time-series-property correlates of heavy-tailed behaviour.

  \item Table.~\ref{tab:sparks} selects populations that were categorized as heavy-tailed and digs into the causes. These are a non-random sample that I was able to verify in the literature.

\end{enumerate}

%- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})

Supporting material to be noted quickly:

\begin{enumerate}

  \item Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).

  \item Time series of all heavy tailed populations (Fig.~\ref{fig:heavy-ts}).

  \item The effect of alternative population dynamics models on $\nu$ estimates (Fig.~\ref{fig:alt}).

  \item Coefficients from modelling covariates of the probability of heavy tails (Fig.~\ref{fig:correlate-coefs}).

  \item Example Stan code for heavy-tailed Gompertz model with AR1 residuals and observation error (a model including everything).

  \item GPDD IDs used.

\end{enumerate}

Modelling result: an increase of 1 time step of data (given that you are ballpark around the mean --- 30 time steps --- to start with) results in an approximately 1\% increase in the expected probability of observing heavy tails. (Using Gelman's `divide by 4' rule for interpreting logistic regression coefficients.)

\section{Discussion}

\begin{enumerate}
  \item Summary of our findings
  \item How do our results mesh with previous related analyses?
  \item Why might we expect to see heavy tails in ecological time-series? (mixture of normals or extreme drivers)
  \item Are the observed frequencies by taxonomic class real or an observational phenomena?
  \item Ways forward
  \item Policy implications
\end{enumerate}

\section{Acknowledgements}

Funding: SCA: SFU Graduate Fellowship, NKD: NSERC SFU \ldots, ABC: \ldots, TAB: \ldots

Earth to Ocean's research group for helpful discussions

Global Population Dynamics Database

Silhouettes: \texttt{phylopic.org}: rabbit by Sarah Werning, grey heron by Ardea cinerea, hoverfly by Gareth Monger. All under Creative Commons Attribution 3.0 Unported license.

Compute Canada's WestGrid high-performance computing resources

\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

\clearpage

\section{Tables}

\LTcapwidth=\textwidth
\bibpunct{}{}{;}{a}{}{;}

\begin{small}
\begin{longtable}{>{\RaggedRight}m{2.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{7.0cm}>{\RaggedRight}p{2.0cm}}

\caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes. Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise. These populations are a non-random sample that we were able to verify in the primary literature.}\\

\toprule
Time series (log scale) & Population & Black swan description & Reference \\
\midrule

\includegraphics[width=2cm]{sparks/6528} &
Shag,
\textit{Phalacrocorax aristotelis},
UK &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
\citep{potts1980}\\

\includegraphics[width=2cm]{sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

\includegraphics[width=2cm]{sparks/7115} &
Fur seal,
\textit{Arctocephalus pusillus},
South Africa &
Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
\citep{shaughnessy1982}\\

\includegraphics[width=2cm]{sparks/10113} &
Willow grouse,
\textit{Lagopus lagopus},
UK &
Parasite and predation effects interacted to cause low years  &
\citep{dobson1995}\\

\includegraphics[width=2cm]{sparks/10162} &
Red grouse,
\textit{Lagopus lagopus scoticus},
UK &
Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
\citep{mackenzie1952}\\

\includegraphics[width=2cm]{sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was submerged in snow caused population crash &
\citep{newton1998} \\

\includegraphics[width=2cm]{sparks/20579} &
Grey heron,
\textit{Ardea cinerea},
UK &
Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
\citep{stafford1971} \\

%\includegraphics[width=2cm]{sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
% &
%\citep{brook2006a}\\

%\includegraphics[width=2cm]{sparks/5019} &
%Barbary macaque,  &
% &
%REF\\

%\includegraphics[width=2cm]{sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
% &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
% &
%\citep{stafford1971}\\

\bottomrule
\label{tab:sparks}
\end{longtable}
\end{small}

% reset citation style:
\bibpunct{(}{)}{;}{a}{}{;}

%\clearpage

\section{Figures}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
  (a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
  Small values of $\nu$ create heavy tails.
  As $\nu$ approaches infinity the distribution approaches the normal distribution.
  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
  (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
  Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
  (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
  Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
  Colour shading behind panels f--h illustrates the region of heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order. Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{correlates.pdf}
\caption{
Potential correlates of heavy-tailed behaviour (indicated by low $\nu$ values). Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
Circles representing four populations are filled with white and represent a shark, crustaceans, and one gastropod.
}
\label{fig:correlates}
\end{center}
\end{figure}

\clearpage

@@

\section{Data selection}

To TODO \ldots

\begin{enumerate}

\item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{"harvest"} as well populations with the words \texttt{"harvest"} or \texttt{"fur"} in the cited reference title.
\item We removed all populations with uneven sampling intervals. I.e\ we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
\item We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
\item Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance. In many cases this was noted for the population, but not in all cases. We inspected each of these  \totalAssumedLog\ time series to make sure our assumption made sense (Fig.~\ref{fig:log10-assumed}).
\item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values. We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fit on.
\item We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}. This assumes that single values of zero result from abundance being low enough that censusing missed present individuals. We turned multiple zero values in a row into \texttt{NA} values. This implies that multiple zero values were either censusing errors or caused by emigration. Regardless, our population  models were fit on a multiplicative (log) scale and so could not account for zero abundance.
\item We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
\item We removed all populations without at least four unique values \citep[following][]{brook2006a}.
\item We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series.  If there were any populations with multiple windows of identical length, we took the most recent window. This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
\end{enumerate}

We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.\ SX). Table~S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

\section{Details on the heavy-tailed Gompertz probability model}

For the Gompertz, the mostly weakly-informative priors I'm using are:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\nu &\sim \mathrm{Exponential}(0.01)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (5).
\end{align*}

See Figure~\ref{fig:priors}.

Priors justification: (TODO this is old)

\begin{itemize}
\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off too far

\item $\lambda$ prior (variance = 100) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X

\item $\nu$ is based on \citet{fernandez1998}; they chose a more informative 0.1 value, we chose a less informative 0.01 and justify it based on performance in supplemental figure X (gives X probability of value less than 10) but constrains the sampling somewhat since above 40 or 50 the shape of the t distribution is almost identical and data are not usually informative at these data quantities... sampler would head off to infinity otherwise

\item $\phi$ given a standard deviation of 1 given prior analyses which suggest autocorrelation in these datasets is minimal; prior is weak enough to allow high or low values, but given little information will stay near our expectation; similar approach in \citep{thorson2014a}

\item $\sigma_\mathrm{proc}$ can be justified based on \citet{gelman2006c} and the expected range of this variable in nature from previous studies \citep[e.g.][]{connors2014}
\end{itemize}

We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. I'm starting with 4 chains and 2000 iterations with the first 1000 as warmup (i.e.\ 4000 total samples). If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again. This continued up to 8000 iterations (16000 total samples) \ldots

\section{Simulation testing the model}

Throughout all of this --- show that the model, if anything, under-predicts heavy tails but with sufficient data is unbiased.

2 parts: how many samples from the true population t distribution do you need to detect low nu? And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?

First part: We drew from t distributions with different nu values and mean of 0, scale of 1. We started with 1600 samples and then fitted again at the first 800, 400, 200,  100, 50, 25. Each time we recorded the nu posterior.

Second part: To generate a series of process deviations with an effective nu approximately equal to the true nu, we generated process deviation sets repeatedly and estimated the mean, scale, and nu values each time. We recorded when the estimated nu was within 0.5 CVs of the true nu and used this set of random seed values in our Gompertz simulation. We then fit AR1 Gompertz models to the simulated datasets with all parameters (except nu) set near the median values estimated in the GPDD.

\section{Alternative population models}

\subsection{Ricker-logistic}

We also fitted a Ricker-logistic model:
\begin{align*}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
\end{align*}
\noindent
where  $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$. The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$. The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$). In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).

To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is less conservative). We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}. We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.

\subsection{Autocorrelated residuals}

We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
\end{align*}
\noindent
In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = 1, \mathrm{max.} = 1)$.

\subsection{Assumed observation error}

We considered a version of the base Gompertz model that allowed for a specified level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}),
\end{align*}

\noindent
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.2$. TODO justify $0.2$.

%\end{spacing}

\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\include{stat-table}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{log10-assumed.pdf} \caption{Populations in which the abundance was specified as or assumed to be recorded as log10 values. The numbers at the front of each panel label are the GPDD IDs.}
    \label{fig:log10-assumed}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{all-clean-ts-3.pdf} \caption{All filtered times series used in our analysis. The abundances are shown on a log10 vertical axis. Colours indicate taxonomic classes. Open black circles indicate abundance values that were interpolated. Closed black circles indicate single abundance values that were recorded as zero but were set to the next lowest observed abundance.}
    \label{fig:all-ts}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Normal}(0, 1)[-1, 1]$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
  \label{fig:priors}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows). The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples. TODO switch to nu = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{sim-gompertz.pdf}
\caption{Simulation testing the Gompertz estimation model when the process error draws are chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
  Dashed horizontal lines show the true population values;
  these true values were chosen to represent approximately the median values as estimated from the GPDD.
  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting (median and 80\% credible intervals).
  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate parameter $\lambda$, and the density dependence parameter $b$.}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{sim-gompertz-boxplots.pdf}
\caption{The distribution of median estimates from the same simulation as described in Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The whiskers extend to $1.5$ times the interquartile range and outliers are shown as dots.}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{check-sim-box.pdf}
\caption{The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy tails) for the same simulation scenarios as shown in Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.}
\label{fig:sim-prob}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{ts-gpdd-heavy-eg-log10-base.pdf}
\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model. Panels are ordered by increasing p($\nu < 10$). Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO add dots for upswings and downswings as in Table 1. The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
\label{fig:heavy-ts}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-comparison.pdf}
\caption{Estimates of $\nu$ from alternative models plotted against the base Gompertz model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. The Gompertz observation error model assumes a fixed standard deviation of observation error of $0.2$ on the log scale.}
\label{fig:alt}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{admb-coefs}
\includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
\caption{Potential correlates of heavy-tailed dynamics.
  (Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length). All covariates were standardized by subtracting their mean and dividing by their standard deviation.
  The beta distribution model (left panel) uses p$(\nu < 10)$ as the response (probability of heavy tails).
  The binomial distribution model (right panel) uses p$(\nu < 10) > 0.5$ vs.\ p$(\nu < 10) \le 0.5$ as the response (heavy vs.\ not heavy tails).
  The models were fitted as mixed effects models with beta or binomial distributions, logit links, and multilevel intercepts for taxonomic class and order nested within class using the \texttt{glmmADMB} \texttt{R} package \citep{fournier2012,glmmadmb}.
  (Bottom panels) Model predictions for the strongest covariate --- time series length --- with all other predictors set to zero (their mean value when standardized). Dots represent observations for individual populations. Lines and shaded regions represent estimates and 50\% and 95\% confidence intervals.
}
    \label{fig:correlate-coefs}
\end{center}
\end{figure}

% \clearpage
%
% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
% \caption{Model predictions. TODO}
%     \label{fig:length-predictions}
% \end{center}
% \end{figure}

\clearpage

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated residuals and a specified level of observation error:

\begin{spacing}{1.15}
\begin{footnotesize}
\begin{verbatim}
data {
  int<lower=3> N;              // number of observations
  vector[N] y;                 // vector to hold ln abundance observations
  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
}
parameters {
  real lambda;                 // Gompertz growth rate parameter
  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
  real<lower=0> sigma_proc;    // process noise scale parameter
  real<lower=2> nu;            // t-distribution degrees of freedom
  real<lower=-1, upper=1> phi; // AR1 parameter
  vector[N] U;                 // unobserved states
  real<lower=0> sigma_obs;     // specified observation error SD
}
transformed parameters {
  vector[N] epsilon;           // error terms
  epsilon[1] <- 0;
  for (i in 2:N) {
    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
                       - (phi * epsilon[i - 1]);
  }
}
model {
  // priors:
  nu ~ exponential(nu_rate);
  lambda ~ normal(0, 10);
  sigma_proc ~ cauchy(0, 2.5);
  phi ~ normal(0, 1);
  // data model:
  for (i in 2:N) {
    U[i] ~ student_t(nu,
                     lambda + b * U[i - 1]
                     + phi * epsilon[i - 1],
                     sigma_proc);
  }
  y ~ normal(U, sigma_obs);
}
\end{verbatim}
\end{footnotesize}

\clearpage

\noindent
The GPDD IDs used in our analysis:

\begin{footnotesize}
\noindent
{\tt
\input{mainids}
}
\end{footnotesize}
\end{spacing}
\end{document}

>>>>

\setlength\parskip{0.16in}
\setlength\parindent{0in}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}

@@

\input{ms}

\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
%\bibliography{/Users/seananderson/Dropbox/tex/jshort,refs}

\clearpage
\section{Tables}

\input{sparks.tex}

\clearpage
\section{Figures}

\input{figs.tex}

@@

\input{supp}

\input{supp-figs}

\end{document}
